---
output: 
  pdf_document:
    keep_tex: yes
    fig_caption: yes
    number_sections: yes
geometry: margin=2cm
title: UV Absorbance characteristics in Northern Lakes
subtitle: https://github.com/rachelbash/absorbance-data-project
author: Rachel Bash
date: April 16th, 2019
abstract: "Absorbance values were measured during a portion of the Cascade Project in the Northern Temperate Lakes LTER network for lakes in northern Wisconsin and the southern part of Michigan's Upper Peninsula. Absorption of light in lakes can be an important variable to study in long-term monitoring programs because of its high correlations between other physical characterstics of lakes, such as dissolved organic carbon content. This study examined the variables that contributed to overall absorbance values and found that depth, dissolved organic carbon, total particulate carbon, and lake were significant predictors that informed the linear model. The study also examined absorbance changes over time in five lakes. East Long Lake and West Long Lake had decreasing absorbance values over time, while Tuesday and Peter Lakes had increasing absorbance values over time. Paul Lake did not display any significant trend over time."
fontsize: 12pt
mainfont: Times New Roman
---



\newpage
\tableofcontents 
\newpage
\listoftables
1           Data Summary .........................................................................6
\newpage
\listoffigures 
\newpage



```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, fig.height=4)
# Set your working directory
getwd()
setwd("/Users/rachelbash/Documents/DUKE/Data Analytics/absorbance-data-project")
# Load your packages
library(tidyverse)
library(dplyr)
library(lubridate)
library(FSA)
library(corrplot)
library(viridis)
library(trend)
library(gridExtra)
library(grid)
library(knitr)
library(bookdown)

# Set your ggplot theme
Rachel_theme <- theme_bw(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right")
theme_set(Rachel_theme)

#import data
carbon.data <- read.csv("./raw-data/NTL-LTER_Lake_Carbon_Raw.csv")
```


# Research Question and Rationale

Absorbance is a unit-less measurement that describes how much a substance can absorb light over a certain range of wavelength. The absorbance values of water samples from lakes can provide details regarding its physical characteristics and the health of the lake. The amount of light entering a lake is a component that drives photosynthesis and lake metabolism (Thrane, 2014). Additionally, lake temperature and its absorbance characteristics are deeply intertwined. With the right equipment, absorbance is fairly easy to measure. Therefore, measuring absorbance in lakes can give researchers insight into other processes happening that depend in part on sunlight, such as algal growth or temperature-dependent biological activities. 

This research project intends to answer three main questions: 

* What contributes to absorbance values in five lakes located in Michigan's Upper Peninsula? 
* Are absorbance values between lakes different?
* Do absorbance values in these five study lakes change over time? 

The data that  answer these questions come from the North Temperate Lakes Project, which seeks to measure data on carbon and other related variables in lakes. My analysis of the data provides a model that shows the variables that best predict absorbance values and also takes a closer look at how absorbance values have changed over time in different lakes. Time variations in absorbance have implications that other physical characteristics are changing, which may damage  biota in the lakes or bring about significant changes in the greater ecosystem that surrounds the lake. Because absorbance value changes can be an indicator, it is important to study this measurement and what contributes to the changes.



\newpage

# Dataset Information

<Information on how the dataset for this analysis were collected, the data contained in the dataset, and any important pieces of information that are relevant to your analyses. This section should contain much of same information as the README file for the dataset but formatted in a way that is more narrative.>

<Add a table that summarizes your data structure. This table can be made in markdown text or inserted as a `kable` function in an R chunk. If the latter, do not include the code used to generate your table.>

The dataset was collected from 1984 to 2016 by researchers working for the Cascade Project and Northern Temperate Lakes Long-Term Ecological Research Network (NTL-LTER) at a total of 14 sites. Samples of water were collected, and then were measured. Measurements included dissolved organic and inorganic carbon, particulate organic matter, partial pressure of carbon dioxide, and absorbance. Absorbance was measured using a spectrophotometer at a wavelength of 440nanometers. 

For some variables, a water depth sample was taken that was measured in meters, while in others, samples were taken to reflect a depth that was proportional across all lakes. Therefore, Hypolimnion, Epilimnion, Metalimnion, and pooled mixed layer (PML) are also included as depth values. All water samples were taken with a syringe and then filtered through a mesh filter in order to remove any large debris or zooplankton.

Data Summary        | Relevant Information
------------------- | ----------------------
Date range          | 1984-06-03 to 2016-08-17
Retrieved from      | NTL - LTER Cascade Project at North Temperate Lakes LTER Core Data Carbon 
Structure           | 15 variables with 13,557 observations
Column variables    | Lake ID, Lake Name, Year, Day No., Date, Depth, Depth ID, TPC, TPN, DIC, PCO2 air, PCO2 water, DOC, Absorbance
Lakes sampled       | Crampton Lake, East Long Lake, Hummingbird Lake, Long Lake, Morris Lake, North Gate Bog, Paul Lake, Peter Lake, Reddington Lake, Roach Lake, Tender Bog, Tuesday Lake, Ward Lake, West Long Lake

```{r, include=FALSE}
#import data
carbon.data <- read.csv("./raw-data/NTL-LTER_Lake_Carbon_Raw.csv")

#set date
carbon.data$sampledate <- as.Date(carbon.data$sampledate, format = "%m/%d/%y")
head(carbon.data$sampledate)

```


\newpage

# Exploratory Data Analysis and Wrangling

<Include R chunks for 5+ lines of summary code (display code and output), 3+ exploratory graphs (display graphs only), and any wrangling you do to your dataset(s).> 

<Include text sections to accompany these R chunks to explain the reasoning behind your workflow, and the rationale for your approach.>

##Importing raw data and identifying its attributes

```{r}

colnames(carbon.data)
str(carbon.data)
summary(carbon.data)
dim(carbon.data)
summary(carbon.data$absorbance)
class(carbon.data$depth)
head(carbon.data$depth, 10)
```

These exploratory commands above function as helpful tools that help me see what kind of shape my data are in. It shows me the size of the data frame, how many NA's I have, what variables I am working with, the classes of my variables, and basic summary statistics. These are important to know and help inform me when making decisions about further analysis. An meaningful detail I discovered while doing the initial exploratory data analysis is that the depth variable has both numeric and factor-level observations, which is why its class is listed as `factor`. In other words, depth was measured in both numeric terms (1 meter, 13 meters, etc), but also in thermally stratified terms, such as Hypolimnion, Metalimnion, and Epilimnion. This was an important discovery that led to further data wrangling and filtering of this specific variable.

##Visualizing the data

As seen by \autoref{fig:foo}, Absorbance values are not normally distributed. This is expected, as we are dealing with ecological data.

```{r foo, echo=FALSE, warning=FALSE, fig.cap="\\label{fig:foo}Absorbance frequency"}
ggplot(carbon.data) +
  geom_histogram(aes(x=absorbance), color="#1f78b4", fill="#1f78b4") +
  labs(x= "Frequency", y="Absorbance value")
```

```{r freqpol, echo=FALSE, warning=FALSE, fig.cap="\\label{fig:freqpol}Absorbance frequency by depth category"}

ggplot(carbon.data) +
  geom_freqpoly(aes(x=absorbance, color = depth)) +
  labs(x= "Absorbance", y = "Frequency", color= "Depth") +
  scale_color_brewer(palette = "Paired")
```


Similarly, \autoref{fig:freqpol} shows that different levels of depth (factor) had difference absorbance frequency values. It was helpful to create this graph to show that absorbance was measured at multiple different water depth levels.


```{r absorbdoc, echo=FALSE, fig.cap="\\label{fig:absorbdoc} Disolved organic carbon and absorbance relationship by year"}
ggplot(carbon.data) +
  geom_point(aes(x = absorbance, y = doc, color=year4), alpha = 0.7) +
  scale_color_distiller(palette="YlGnBu", direction = 1) +
  labs(x="Absorbance", y="Dissolved Organic Carbon, mg/L", color="Year")
```

 \autoref{fig:absorbdoc} shows a positive relationship between dissolved organic carbon and absorbance, with a layer of color by year. This result is expected, and it gave me a good sense of what to expect during my analysis portion of the project. It is interesting to note that as time went on, measures for both absorbance and for DOC began to shrink to smaller values, as seen with the color gradient by year. Another thing this plot tells me is that absorbance probably wasn't measured in the early times of data collection, as there are no points before 1990, as indicated by the yellow color on the graph.
 
##Data Wrangling
```{r}
carbon.data.processed <- carbon.data %>%
  filter(depth %in% c("PML", "Hypolimnion", "Epilimnion", "Metalimnion")) %>%
  filter(lakename %in% c("Peter Lake", "Paul Lake", "East Long Lake", "Tuesday Lake", "West Long Lake")) %>%
  select(lakename:depth_id, DIC_mg, doc, absorbance, tpc)

```
 
There were many things to consider when wrangling my data to a more manageable and workable dataset. I noticed that all absorbance values had associated depth measurements using only the thermally stratified depth categories. Therefore, I filtered out any depth that was measured in meters, in order to simplify the process. Next, I chose the five lakes in the dataset that had the most number of data points. Shortening the lake list from 14 to 5 gives the research project a more focused view and potentially stronger relationships among variables. Lastly, I selected only the columns that I wanted to study and that could be analyzed in relation to absorbance values. These variables included lake name, depth, dissolved inorganic carbon, dissolved organic carbon, total particulate carbon, and absorbance.

##Correlation between continuous variables
```{r corr, echo=FALSE, fig.cap="\\label{fig:corr} Correlation plot between continuous variables"}
carbon.data.processed.naomit <- carbon.data.processed %>% 
  select(DIC_mg:tpc) %>%
  na.omit()

carbonCorr <- cor(carbon.data.processed.naomit)
corrplot.mixed(carbonCorr, upper = "ellipse")
```

The last piece of data exploration I completed was visualizing the correlation between the continuous variables in the data. \autoref{fig:corr} illustrates the relationships between each of the continuous variables in question. All relationship correlations range pretty low to moderate, with DOC and absorbance having the lowest correlation coefficient of -0.12, and DIC and DOC having the highest negative correlation coefficient of -0.58. It is important to consider this visualization critically, as the data have been thoroughly reduced at this point, leaving much fewer data points than what we started with, which could manipulate the strength (or weakness) of these correlation coefficients.


\newpage

# Analysis
<Include R chunks for 3+ statistical tests (display code and output) and 3+ final visualization graphs (display graphs only).>

<Include text sections to accompany these R chunks to explain the reasoning behind your workflow, rationale for your approach, and the justification of meeting or failing to meet assumptions of tests.>

##Differences in absorbance values across lakes

It was important for me to know whether absorbance values were significantly different across the five lakes of interest. This can be answered by a simple ANOVA test. However, the data must meet certain criteria. Fist, data had to be normally distributed, and second, equal variance across groups must exist. I tested these assumptions using the Shapiro Wilk test and the Bartlett test, respectively. Both tests resulted in significant p-values, indicating that the data fail the tests for normality and equal variances. 

Therefore, another method had to be utilized. I opted for a non-parametric test called the Kruskal Wallis test, a great alternative to ANOVAs. Here, I received a significant p-value result, indicating that there is a significant difference in absorbance values across different lakes (chi-squared = 739.62, df = 4, p-value < 2.2e-16). A non-parametric post-hoc test (Dunn Test) reveals that all lakes' mean absorbances values are significantly different from all other lakes (p-values < 0.05).  \autoref{fig:vio} illustrates how absorbance values vary vastly by lake. Even though all of these lakes are located close to each other in Michigan's Upper Peninsula along the Wisconsin border, it is clear that absorbance values can vary greatly among them. Even Peter and Paul Lakes, whose mean absorbance values do look fairly close, do possess a significant p-value in the post-hoc test, indicating that they are statistically significantly different from one another.


```{r}
# test for normality
shapiro.test(carbon.data.processed$absorbance
             [carbon.data.processed$lakename == "Tuesday Lake"])
shapiro.test(carbon.data.processed$absorbance
             [carbon.data.processed$lakename == "Paul Lake"])
shapiro.test(carbon.data.processed$absorbance
             [carbon.data.processed$lakename == "Peter Lake"])
shapiro.test(carbon.data.processed$absorbance
             [carbon.data.processed$lakename == "East Long Lake"])
shapiro.test(carbon.data.processed$absorbance
             [carbon.data.processed$lakename == "West Long Lake"])
#result: all have significant p-values, meaning they are not normally distributed data 

#bartlett test to determine whether there is equal variance between groups
bartlett.test(carbon.data.processed$absorbance ~ carbon.data.processed$lakename) 
#result: significant p-value, not equal variances

#non-parametric test instead
kruskal.test(carbon.data.processed$absorbance ~ carbon.data.processed$lakename)
#lakename is a significant predictor of absorbance

#post-hoc non-parametric test
dunnTest(carbon.data.processed$absorbance ~ carbon.data.processed$lakename) 
#shows all lakes differ from one another significantly 

#created correct figure caption and auto
```


 
```{r vio, echo=FALSE, fig.cap="\\label{fig:vio}Absorbance differences by lake"}
ggplot(carbon.data.processed, aes(x=lakename, y = absorbance, color=lakename)) +
  geom_violin(draw_quantiles = 0.5) +
  labs(x="Lake Names", y="Absorbance Distribution") +
  theme(axis.text.x = element_text(angle = 45,  hjust = 1), axis.text = element_text(size=10),  legend.title = element_blank(), legend.position="none") +
  scale_color_manual(values = c("#7fcdbb", "#41b6c4", "#1d91c0", "#225ea8", "#0c2c84"), 
                   breaks = c("East Long Lake", "Peter Lake", "Paul Lake", "Tuesday Lake", "West Long Lake"))
```
\newpage

##Linear regression model

In order to determine what factors contribute to absorbance values, I conducted a step-wise linear regression model. I was not able to include DIC in the full model, because DIC was only measured at numeric depths and not thermally stratified depths. Therefore, no data between absorbance and DIC overlapped. The full model included the following parameters: depth, DOC, Lake Name, and Total Particulate Carbon (TPC). 

By performing a step-wise linear regression model, I can use the lowest Akaike's Information Criterion (AIC) value to determine the ideal statistical model that balances both simplicity and statistical power. My regression analysis showed that all variables are significant and allow us to best predict absorbance values. The resulting linear expression is as follows:

$$Absorbance = 0.18(Epi*East) + 0.17(Hypo) + 0.02(DOC) -$$ 
$$0.18(Paul)  - 0.24(Peter) - 0.23(Tuesday) - 0.19(West) + 0.02(TPC) $$
For example, for every one unit increase in DOC, absorbance value increases by 0.02 units. Alternatively, if absorbance is measured in Peter Lake, absorbance will decrease by 0.24 units. The step function shows that all variables, except for Metalimnion and PML depth categories were significant predictors of absorbance, and the full model had an adjusted R-squared value of 0.86, which is quite significant. However, it is important to note that because of missing values within each of the variables, there is an exceptionally high degrees of freedom value of 1119. 

  \autoref{fig:lin} illustrates all predictors, with the exception of lake name, and their relationship to absorbance. 
 
```{r}
#Step-wise linear regression

steplm <- lm(data=carbon.data.processed, absorbance ~ depth + doc  + lakename + tpc)
step(steplm) 
#taking out none would result in the lowest AIC value

#full-model is the best, as shown by the step 
fullmodel <- lm(data=carbon.data.processed, absorbance ~ depth + doc  + lakename + tpc)
summary(fullmodel)
```

```{r lin, echo=FALSE, fig.cap="\\label{fig:lin} Facet plot of absorbance by TPC, DOC, and Depth"}

ggplot(carbon.data.processed, aes(x=tpc, y=absorbance)) +
         geom_point(aes(color=doc), alpha = 0.5) +
         facet_wrap(vars(depth)) +
         theme(axis.text.x = element_text(angle = 45, hjust = 1), strip.text.x=element_text(size = 9)) +
         scale_color_viridis(option="viridis", direction = -1) +
         xlim(0,5) +
  labs(x = "Total Particulate Carbon (mg/L)", y = "Absorbance", color = "DOC")

```
\newpage


##Time Series Analysis

A time series analysis will show whether absorbance values in lakes have changed over time. I used a Mann Kendall test to determine whether there is a monotonic overall trend in absorbance over time for the five lakes of interest. While there is perhaps seasonality differences in absorbance values, I am only interested in yearly trends, and so have decided not to look at seasonal differences of absorbance.

To complete a time series analysis, I decided to split up the data by lake so that I was able to determine whether there was a trend in each lake. I also decided to filter the data by only observing absorbance values at the Hypolimnion, which is the stratified column below the thermocline at the bottom of the lake. I thought that choosing one depth height would be appropriate, and I chose the bottom because I thought it would be the most uniform layer among all lakes.

East Long Lake had a significant Mann Kendall test result, which indicated a signicant negative monotonic trend over time (p-value = 2e-9). Using a series of pettitt tests and additional Mann Kendall tests, I discovered three change points in East Long Lake's absorbance data as well. West Long Lake also had significant negative trend from the beginning to the end of the data, with one change point detected (p-value = 1.18e-06). Peter and Tuesday Lakes both had a significant over positive trend over time, meaning that the Mann Kendall test shows that absorbance values in these lakes increased over time (p-value Peter = 1.47e-06; p-value Tuesday= 0.0005). Both lakes also had two change points. Lastly, Paul Lake's Mann Kendall test produced a non-significant p-value, suggesting that there is no significant monotonic trend in absorbance values over time (p-value = 0.10). However, a change point was still detected in data. \autoref{fig: time} summarizes and visualizes the findings.

```{r}
#trimming data to only look at points of interest
carbon.data.processed.trimmed <- carbon.data.processed %>%
  filter(depth == "Hypolimnion") %>%
  select(absorbance, sampledate, lakename) %>%
  filter(sampledate > as.Date("1996-06-01") & sampledate < as.Date("2016-08-17"))

```


### East Long Lake

```{r}
#East Lake
East.mktest <- filter(carbon.data.processed.trimmed, lakename == "East Long Lake")

#run MK test
mk.test(East.mktest$absorbance) #p=2e-9, so there is a significant negative trend from beginning to end of data

# Test for change point
pettitt.test(East.mktest$absorbance) #change point detected at place 30 1998-05-28

#run second MK test for each change point range
mk.test(East.mktest$absorbance[1:29]) #p=1e-5, significant positive trend in this range
#test for change point
pettitt.test(East.mktest$absorbance[1:29]) #change point detected at place 17 1997-06-12
mk.test(East.mktest$absorbance[1:16]) #no trend
mk.test(East.mktest$absorbance[17:29]) #no trend

mk.test(East.mktest$absorbance[30:73]) #significant negative trend over time
pettitt.test(East.mktest$absorbance[30:73]) #change point detected at place 17 (which means place 47 - 1999-06-09)
mk.test(East.mktest$absorbance[30:46]) #no trend
mk.test(East.mktest$absorbance[47:73]) #no trend


```

###West Long Lake

```{r}
#split data by lake
West.mktest <- filter(carbon.data.processed.trimmed, lakename == "West Long Lake")

#run MK test
mk.test(West.mktest$absorbance) #p is significant, so there is a significant negative trend from beginning to end of data

# Test for change point
pettitt.test(West.mktest$absorbance) #change point detected at place 44 1998-08-31

#run second MK test for each change point range
mk.test(West.mktest$absorbance[1:43]) #no trend

mk.test(West.mktest$absorbance[44:87]) #no significant trend in this range

  
```


###Peter Lake

```{r}
#Peter Lake
Peter.mktest <- filter(carbon.data.processed.trimmed, lakename == "Peter Lake") %>% na.omit()

#run MK test
mk.test(Peter.mktest$absorbance) #p val is significant so there is a significant positive trend from beginning to end of data

# Test for change point
pettitt.test(Peter.mktest$absorbance) #change point detected at place 202 2011-06-06

#run second MK test for each change point range
mk.test(Peter.mktest$absorbance[1:201]) #p>0.05, no significant trend in this date range

mk.test(Peter.mktest$absorbance[202:287]) #significant positive trend over time
pettitt.test(Peter.mktest$absorbance[202:287]) #change point detected at place 50 (which means place 50+202=252 - 2014-06-30)
mk.test(Peter.mktest$absorbance[202:251]) #no trend
mk.test(Peter.mktest$absorbance[252:287]) #no trend

```


###Paul Lake

```{r}
#Paul Lake
Paul.mktest <- filter(carbon.data.processed.trimmed, lakename == "Paul Lake") %>% na.omit()

#run MK test
mk.test(Paul.mktest$absorbance) #p>0.05 so no significant over time

# Test for change point
pettitt.test(Paul.mktest$absorbance) #change point detected at place 72 2001-08-20

#run second MK test for each change point range
mk.test(Paul.mktest$absorbance[1:71]) #no trend

mk.test(Paul.mktest$absorbance[72:291]) #no trend


```

###Tuesday Lake


```{r}
#Tuesday
Tuesday.mktest <- filter(carbon.data.processed.trimmed, lakename == "Tuesday Lake") %>% na.omit()

#run MK test
mk.test(Tuesday.mktest$absorbance) #p is significant, so there is a significant positive trend from beginning to end of data

# Test for change point
pettitt.test(Tuesday.mktest$absorbance) #change point detected at place 68 2014-06-25

#run second MK test for each change point range
mk.test(Tuesday.mktest$absorbance[1:67]) #p<0.05, significant negative trend in this range
#test for change point
pettitt.test(Tuesday.mktest$absorbance[1:67]) #change point detected at place 44 2012-08-22
mk.test(Tuesday.mktest$absorbance[1:43]) #no trend
mk.test(Tuesday.mktest$absorbance[44:67]) #no trend

mk.test(Tuesday.mktest$absorbance[68:102]) #no trend


```


```{r time, echo=FALSE, fig.cap="\\label{fig:time} Absorbance in East Long Lake, West Long Lake, Peter Lake, Paul Lake, and Tuesday Lake over time", fig.height=7, fig.width=9}
#East
East <- ggplot(East.mktest, aes(x=sampledate, y=absorbance)) +
  geom_point(color="#7fcdbb") +
  geom_vline(xintercept=as.Date("1997-06-12")) +
  geom_vline(xintercept=as.Date("1999-06-09")) +
  geom_vline(xintercept = as.Date("1998-05-28")) +
  geom_smooth(method=lm, color="black") +
  theme(axis.title.x=element_blank(), axis.title.y=element_blank(), plot.title=element_text(hjust=0.5)) +
  labs(title="East Long Lake")


#West
West <- ggplot(West.mktest, aes(x=sampledate, y=absorbance)) +
  geom_point(color = "#0c2c84") +
  geom_vline(xintercept=as.Date("1998-08-31")) +
  geom_smooth(method=lm, color = "black") +
  theme(axis.title.x=element_blank(), axis.title.y=element_blank(), plot.title=element_text(hjust=0.5)) +
  labs(title="West Long Lake")
  

#peter
Peter <- ggplot(Peter.mktest, aes(x=sampledate, y=absorbance)) +
  geom_point(color="#1d91c0") +
  geom_vline(xintercept=as.Date("2014-06-30")) +
  geom_vline(xintercept=as.Date("2011-06-06")) +
  geom_smooth(method=lm, color="red") +
  theme(axis.title.x=element_blank(), axis.title.y=element_blank(), plot.title=element_text(hjust=0.5)) +
  labs(title="Peter Lake")


#paul
Paul <- ggplot(Paul.mktest, aes(x=sampledate, y=absorbance)) +
  geom_point(color="#41b6c4") +
  geom_vline(xintercept=as.Date("2001-08-20")) +
  geom_smooth(method=lm, color="black") +
  theme(axis.title.x=element_blank(), axis.title.y=element_blank(), plot.title=element_text(hjust=0.5)) +
  labs(title="Paul Lake*")


#Tuesday
Tuesday <- ggplot(Tuesday.mktest, aes(x=sampledate, y=absorbance)) +
  geom_point(color="#225ea8") +
  geom_vline(xintercept=as.Date("2014-06-25")) +
  geom_vline(xintercept=as.Date("2012-08-22")) +
  geom_smooth(method=lm, color="red") +
  theme(axis.title.x=element_blank(), axis.title.y=element_blank(), plot.title=element_text(hjust=0.5)) +
  labs(title="Tuesday Lake")


#ALL
All <- ggplot(carbon.data.processed.trimmed, aes(x=sampledate, y = absorbance, color=lakename)) +
  geom_point(aes(color=lakename), size = 2) + theme(legend.position="top", legend.title = element_blank(), legend.text=element_text(size=10)) +
  theme(axis.title.x=element_blank(), axis.title.y=element_blank()) +
  scale_color_manual(values = c("#7fcdbb", "#41b6c4", "#1d91c0", "#225ea8", "#0c2c84"), 
                   breaks = c("East Long Lake", "Peter Lake", "Paul Lake", "Tuesday Lake", "West Long Lake"), 
                   labels = c("East Long Lake", "Peter Lake", "Paul Lake", "Tuesday Lake", "West Long Lake"))
#All

#PUT EVERYTHING TOGETHER!
Absorbance.figure <- grid.arrange(All, East, West, Tuesday, Peter, Paul, ncol=3, nrow=3, 
             layout_matrix = rbind(c(1, 1, 2),
                                   c(1, 1, 3),
                                   c(4, 5, 6)), 
            top=textGrob("Absorbance over Time", gp=gpar(fontsize=20)), bottom = "Date", 
            left = "Absorbance")
Absorbance.figure

```


\newpage

# Summary and Conclusions
<Summarize your major findings from your analyses. What conclusions do you draw from your findings? Make sure to apply this to a broader application for the research question you have answered.>

Absorbance is able to be predicted by lake, depth, total particulate carbon, and dissolved organic carbon concentration. The model that included all of these variables were able to explain 85% of the variation of absorbance within the data. Absorbance values are also significantly different in all five lakes examined, as determined by the Kruskal Wallis and Dunn Tests. 

It is important to note that while these results are desirable and seem to explain a lot of the variation, there are many other factors that can contribute to absorbance values that were not measured in this dataset. Color of the lake, total depth of the lake, phytoplankton and its pigments, organic materials that were filtered out, and many other potential contributors were not considered. While depth was a variable, only the depth category of Hypolimnion was considered, which did not take into account the total depth of the water in the lake. Absorbance can also be measured at different wavelengths to determine the peak absorbance curve for different substances. The absorption of light also affects the amount of total energy captured by a substance. 

Absorbance characteristics of lakes can inform researchers of valuable information about other characteristics of the lakes that may not be measureable (Beaucler, 2001). Because absorbance is relatively easy to measure with a spectrophotometer, absorbance spectroscopy can be used to characterize and predict other characteristics. It can be an indication of the aesthetic quality of lakes,  amount of DOC or algae or other substances in lakes, retention time, or even latitude (Erlandsson et al., 2012). 

Absorption of light in lakes is a great phenomenon that is easy to study and gives information about many other aspects of the lake's ecosystem. Understanding the relationship between absorbance, depth of the water, DOC content, and TPC content gives us a clearer understanding of the biology and physical characteristics of the five examined lakes in this larger dataset from the Northern Temperate Lakes project.



\newpage

# References

Beaucler, K. B., & Gunn, J. M. (2001). Ultraviolet absorbance in lakes near the metal smelters in Sudbury, Canada. Journal of Environmental Monitoring: JEM, 3(6), 575–579.

Erlandsson, Martin & Futter, Martyn & Kothawala, D & Köhler, S. (2012). Variability in spectral absorbance metrics across boreal lake waters. Journal of environmental monitoring : JEM. 14. 2643-52. 10.1039/c2em30266g.

Thrane, JE., Hessen, D.O. & Andersen, T. Ecosystems (2014) 17: 1040. https://doi.org/10.1007/s10021-014-9776-2

 

